
\section{Introduction}

% Motivation

% more on why programming languages needs to have a way to model these first class time-varying values. 
% then btw, rizzo does this... singals... updating in place... be careful with this ...
Functional reactive programming (FRP) is a programming paradigm for building \textit{reactive} programs in a declarative and functional style. FRP languages model time-varying values as first-class values called \textit{signals}. 
%A sentence or two on the problems/issues that arise in FRP
However, without proper restrictions this style of programming can easily lead to programs that are non-causal, non-productive and contain space leaks.

The FRP language Rizzo~\cite{rizzo_modal_2025} solves these issues using modal types and operational semantics for updating signals in place.
Rizzo maintains a global heap of signals from which all signals are updated when an input is received from the environment.
This process may create intermediate signals which serve no purpose post update and should therefore be erased early to avoid superfluous work. This makes memory management particularly important for Rizzo. 
%A sentence or two to bridge the gap - why we need RC
Reference counting is a strategy for managing dynamic memory that allows immediate erasure of unused heap objects, this property is ideal for Rizzo.

% Goals of the thesis
In this thesis we introduce a compiler for Rizzo that automatically inserts the necessary reference-counting instructions. The compiler outputs C code and includes a library for the reactive semantics of Rizzo.
We base our implementation on the work of \citeauthor{ullrich_counting_2021}~\cite{ullrich_counting_2021} which presents an algorithm for reference counting functional programming languages with memory reuse for \textit{functional-but-in-place} programming.

The results of our thesis show \dots

% In this thesis we introduce a compiler from Rizzo code to a reference counted intermediate representation. We then include a transpiler from this intermediate representation to C code, with a library for the advance and update semantics of Rizzo.
% \dots
% We have based the reference counted intermediate representation on the \citetitle{ullrich_counting_2021} by \citename{ullrich_counting_2021}{author} \cite{ullrich_counting_2021}, which is a technique for reference counting in functional languages. We have adapted this technique to work with the semantics of Rizzo, \dots we describe the details of this in 
