\section{Automatically reference counting Rizzo}

\todo{Quick refresh of assumptions from \textit{Reference Counting section}}

The algorithms created by \citeauthor{ullrich_counting_2021}~\cite{ullrich_counting_2021} are defined on the functional intermediate representations (calculi) $\lambda_{pure}$ and $\lambda_{RC}$ where the latter is an extension of the first with instructions for reference counting.
\todo{Figure shows the language \dots}. 
The \texttt{pap} keyword explicitly marks partial applications, $\texttt{ctor}_i \; \bar y$ is an application of the $i\text{-th}$ constructor of some erased type, and $\texttt{proj}_i \; x$ reads the $i\text{-th}$ field of a constructor value $x$. Partial and full function application may only happen on the names of global functions also called constants.

\[
\begin{aligned}
	w, x, y, z&\in Var\\
    l &\in \textcolor{greenadditions}{Literal \qquad ::= \mathbb{Z} \mid \text{true} \mid \text{false} \mid \text{never} \mid string}\\
    p &\in \textcolor{greenadditions}{Primitive \qquad ::= x\mid l}\\
	c &\in Const \\
	e &\in Expr \qquad ::= c\,\textcolor{greenadditions}{\bar p} \mid \mathbf{pap}\; c\,\textcolor{greenadditions}{\bar p} \mid x\; p \mid \mathbf{ctor}_i\,\textcolor{greenadditions}{\bar p} \mid \mathbf{proj}_i\; x \\
	F &\in FnBody \; ::= \mathbf{ret}\; \textcolor{greenadditions}{p} \mid \mathbf{let}\; x = e;\; F \mid \mathbf{case}\; x\; \mathbf{of}\; \textcolor{greenadditions}{\overline{\langle \mathbb{Z}, F \rangle}} \\
	f &\in Fn \qquad ::= \lambda\,\bar y.\; F \\
	\delta &\in Program \; = Const \to Fn
\end{aligned}
\]

Another important observation is that these languages are defined in the style of A-normal form \todo{cite flanagan}, where expressions are non-recursive and are only allowed on the right-hand-side of let-bindings. 
In our representation, however, this constraint is relaxed by allowing variable names and literals (i.e.~integers, booleans, and strings) in $\texttt{ret}$ and as arguments for function and constructor application. This makes translation easier and reduces program size by allowing $\texttt{ret}\;20$ instead of requiring $\texttt{let} \; x = 20; \; \texttt{ret} \; x$.
Our representation additionally adds the specialised $\texttt{Signal}$ constructor which can be used in place of $\texttt{ctor}_i$. This distinction is necessary for the runtime to know when a memory cell should go into the signal heap (see Section~\ref{section:frp:signal_heap}).~\todo{We have also added top-level values!}

\subsection{Transformations}

Our strategy for inserting reference count instructions is to first transform the Rizzo syntax tree into our modified $\lambda_{pure}$ and then run the $R$, $collect_{\mathbb{O}}$, and $C$ in that order. 
To do so, we must run the Rizzo AST through the transformations shown in \todo{Figure xx} because A-normal form is not the only restriction imposed by the intermediate language, they also impose a handful of restrictions as described in Section~\ref{sec:background:beans:assumptions}.

\todo{Here comes a pretty little flow chart showing how we chain these transformatiosn together, and a little explanation of why/what in relation the assumptions}

\subsection{Counting Rizzo's beans}

\todo{Walk through the 3 steps and explain: Purpose and how have we adapted it for Rizzo?}



\todo{Our adaptation of the immutable bean algorithm. Our modifications. How do we insert these refcount-instructions}

\todo{How do we capture/model the Rizzo constructors, i.e. Delay, Later, Signal, \dots}
