\section{Limitations}\label{sec:limitations}

\subsection{Reuse of Rizzo beans}

As we described in Section~\ref{section:refcount:beans:counting}, the reactive semantics limit reuse in signal combinators to non-recursive combinators such as $stop$. In recursive combinators, the \textbf{reset} instructions are still inserted even though they will never succeed and these then force the insertion of additional \textbf{inc} instructions. 
The reason is that the reference counting algorithms always insert \textbf{reset}/\textbf{reuse} pairs whenever they encounter a suitable constructor expression, even if the memory will be needed again later.
A natural way to fix this would be to not insert \textbf{reset}/\textbf{reuse} in branches where the tail bound by the match is used in a recursive call. Preventing inserts at the level of branches would still allow reuse in combinators like $stop$ that have non-recursive and recursive branches. 

We are not the first to observe that the eager \textbf{reset}/\textbf{reuse} insertion can cause problems. 
\citeauthor*{frame_limited_reuse}~\cite{frame_limited_reuse} 
present a program similar to our $map$ combinator, but their concern is instead that the resulting code can hold onto memory for too long.
Their solution is to reverse the order in which \textbf{reset}/\textbf{reuse} instructions and \textbf{inc}/\textbf{dec} are inserted, so that \textbf{reset} instructions are only inserted when they can replace a \textbf{dec}.
This would also solve our issue of redundant and always failing \textbf{reset} instructions. 

\subsection{The Later applications forgotten to time}\label{sec:later_applications_forget}

\texttt{Later applications} (\texttt{LaterApp}) call arbitrary functions and are thus not restricted by the semantics of the runtime to always return the same value. This means, if multiple signals depend on the same Later application in a single time step, that application will be evaluated once for each dependency rather than sharing the result. This is inefficient both in terms of time and memory, as the Later application may be expensive to compute and will grow the heap linearly with the number of dependencies on the Later value.

In Section~\ref{section:eval:heap_behaviour}, the signals $4$ and $6$ arise from exactly this situation: the same later application is evaluated twice, once for each dependency. Ideally, the signal $4$ would have been cached and returned for the second dependency, which would have prevented the creation of signal $6$ and the associated memory allocation. 

A current workaround is to temporarily turn the affected later computation into a signal, so that the result is cached in the signal heap:

\begin{verbatim}
    tail ("" :: mk_sig (wait console))
\end{verbatim}

For us to use the \texttt{tail} we need to turn the \texttt{Later (Signal String)} from \texttt{mk\_sig} into a \texttt{Signal String}. To do this, we take the \texttt{LaterApp} from \texttt{mk\_sig}, and apply it to the signal constructor (\texttt{::}) with a dummy head value (in this case the empty string). This instantiates a signal in the current time step, which we can use the \texttt{tail} constructor on to effectively cache the result of the later application. 
The trade-off here is we abandoned the lazy creation of the signal, and instead create the signal on initialisation. This means that if the signal is never used, we have unnecessarily created a signal and allocated memory for it. However, if the signal is used multiple times, we save on the redundant later applications and the associated memory allocations, which can be a significant improvement in performance and memory usage.

A proper fix could require us to change the LaterApp type to contain 2 additional fields. One field for the cached value and one for a flag to indicate which time step the cached value is valid for. The evaluation of the Later value would then first check if the cached value is valid for the current time step and return it if it is, and otherwise compute the Later value, store it in the cache, and return it.

% A similar problem can be observed for the Delay types, which are also just functions. However, the Delay type is less likely to be used in a way that causes multiple dependencies on the same Delay value, as it is more common to use the Delay type for delaying a signal by one time step, which would not cause multiple dependencies on the same Delay value within the same time step. 
