% \todo{This section should prepare the reader for why we have structured the heap the way we have and why we have chosen to add a signal constructor to the AST.
% We do this by first listing all the semantics we need to adhere to, and how we have done so (this also forms the base of why we have the memory structure we have chosen).}

For our language implementation to work we must combine the semantics of Rizzo~\cite{rizzo_modal_2025} with the reference counting semantics of \citetitle{ullrich_counting_2021}~\cite{ullrich_counting_2021}. This section aims to provide a description of how these two worlds merge to become \todo{RizzoLang}, rather than a comprehensive explanation of each. 
To do so, we first provide an abstract description of the semantics, then give a more concrete account of how this is implemented in the runtime of the language, and finally how reference counting is fitted in.

% Therefore, we focus on the necessary adjustments and the unavoidable parts such as the signal heap.

% Our implementation depends on adhering to the reactive semantics of Rizzo as defined in \citetitle{rizzo_modal_2025}~\cite{rizzo_modal_2025}, while also implementing the reference counting semantics from \citetitle{ullrich_counting_2021}~\cite{ullrich_counting_2021}. In this section we describe how we have structured our implementation to meet these requirements, and how the design of the heap and AST supports these semantics.

\subsubsection{Semantics: Rizzo}

In Rizzo~\cite{rizzo_modal_2025} the operational semantics is split in two: The evaluation semantics for expressions, and the reactive semantics that describe how Rizzo programs react when time is stepped forward.

The evaluation semantics follow a standard eager evaluation strategy. The only exception being $\text{delay} \; e$ for which the argument $e$ is lazily evaluated.
% We have slightly different evaluation semantics because we allow general recursion.
We change the semantics presented by \citeauthor*{rizzo_modal_2025}~\cite{rizzo_modal_2025} \todo{only a little}. We exclude the $\textbf{fix}\text{(-point)}$ construct which is used to guarantee productivity by guarding recursion with the \texttt{Delayed} type. That is, our implementation makes no guarantees about productivity, a trade-off we are willing to make given the usefulness of general recursion in functional programming.

%Trying to say something like: Rizzo~\cite splits the reactive semantics in three: *which? What is their purpose*. We'll point our focus to the update semantics, since this is the most important for the problem at hand
The reactive semantics define the reactive behaviour of Rizzo, i.e.\ what happens when any channel has produced a value and time is moved forward. 
These semantics are split in three parts:
The \textit{advance semantics} which describe how a $\texttt{Later}\, A$ value is advanced to produce a value of type $A$; 
the \textit{update semantics} which, using the advance semantics, describe how a signal is updated and brought forward in time;
and the \textit{step semantics} which describe the process of applying the update semantics to every signal when time is \textit{stepped} forward.

%Goal: Goal is to explain the update semantics. Explain the now vs. earlier heap.
For the purpose of memory management, we will focus our attention on the update semantics but first we take a detour to understand the conceptual structure of the heap.
In the typical way, the heap holds onto all dynamically allocated data such as tuples, signals, function closures, etc.
Among these, signals are special because they must be traversed as part of the reactive semantics. Rizzo therefore distinguishes between the \textit{heap}, storing all dynamically allocated data, and the \textit{signal heap}, a subset of the heap containing only signals. This signal heap is further divided in two parts: the \texttt{now heap}, storing only signals that belong to the current time step; and the \texttt{earlier heap} which holds all the stale signals to be updated.

%Goal: explain update semantics
At runtime when some channel $\mathbf{C}$ produces a value, the step semantics describe that all signals must be moved to the \texttt{earlier heap}, marking their data as stale.
Then as described by the update semantics, each signal is moved one-by-one from the \texttt{earlier heap} to the \texttt{now heap}. To move a signal we first check if its tail is dependent on $\mathbf{C}$ and if so, the tail is \textit{advanced} to obtain a new signal whose data is copied into the original signal and the original signal is moved to the the \texttt{now heap}. Otherwise, the signal is moved to the \texttt{now heap} unchanged.
%We avoid talking about the updated flag - that is an implementation detail. Belongs to advance

%Goal: Here comes a digest, what does all of the above mean? 
However, since chains of signals built using combinators form data-dependencies the move-ordering is significant. 
Boiled down, the reactive semantics require that:
\begin{itemize}
    \item Any new signal must only depend on other signals in the \texttt{now heap}.
    % \item When an input is produced on a channel, all signals with a tail that depends on the clock of that channel must to be updated.
    \item Signals must be updated based on a topological order of their dependencies. If signal B depends on signal A, signal A must appear before signal B in the heap and thus be updated first.
    % \item Evaluating the tail of a signal produces a new signal, which must be copied back into the original signal. %This means that the original signal is updated in place, and any references to it remain valid.
    \item If no reference exists to the intermediate signal produced by evaluating the tail, then the intermediate signal must be deallocated to prevent duplicate work.
    \item New signals must be inserted into the \texttt{now heap}, to prevent them from being evaluated in the current time step.
    % \item 
\end{itemize}

%Goal: Okay, so now we know what is required of the heap structure. How do we represent it? -> Linked list, advantages thereof.

Based on these requirements, we have chosen to structure our heap as a linked list. Using a linked list has the advantage of:
% This gives us a simple way to maintain the topological order of signals, it allows for efficient insertion and removal of signals, 
%  as we can ensure that when a new signal is created it is added after the split between the \texttt{Now} and \texttt{Earlier} heaps. \dots

\begin{itemize}
    \item Representing the \texttt{now} and \texttt{earlier} heaps in one data structure, and expressing a split between them using a pointer (cursor).
    \item Moving signals from the \texttt{now} heap to the \texttt{earlier} heap in O(1) time by resetting the cursor to the start of the heap. Likewise moving signals back to the \texttt{now} heap by moving the cursor forward.
    \item O(1) insertion that maintains the topological order of signals by adding them in the end of the \texttt{now} heap, we ensure that any signal can only depend on signals that were created before it.
    \item O(1) removal of signals when their reference count drops to zero.
    \item Stable references to signals, as insertions and removals do not move other signals in memory, so pointers to signals remain valid throughout their lifetime.
\end{itemize}

\subsubsection{Piecing it all together}

The evaluation semantics of Rizzo~\cite{rizzo_modal_2025} and the operational semantics of \citeauthor*{ullrich_counting_2021}~\cite{ullrich_counting_2021} are both built on the same core: an eagerly evaluated lambda calculus. This common structure allows us to reuse the semantics of \citeauthor*{ullrich_counting_2021}~\cite{ullrich_counting_2021} for most of Rizzo, including both the evaluation rules and the reference counting.

%GOAL: We should mention that all constructors are heap allocated.
In the semantics of \citeauthor*{ullrich_counting_2021}~\cite{ullrich_counting_2021}, there are only two kinds of heap-allocated values: function closures, which represent partial applications, and constructor values of the shape $\textbf{ctor}_i$ which represent values created using the $i\text{-th}$ constructor of a tagged union.
This is already enough to model most of Rizzo's data constructors. 
%GOAL: How do we model Later, Delayed, Sync, etc.?
In particular, values of the \texttt{Sync}, \texttt{Later}, and \texttt{Delayed} types can be represented as regular tagged unions, where each constructor is assigned its own tag.

%GOAL: EXCEPT signals, for those we extend the semantics with a specialised signal constructor, because ...
The exception is the \texttt{Signal} type. Although signals, like other constructors, are just containers of values, they must be treated differently because both their creation and erasure modify the signal heap.
We therefore introduce a specialised \textbf{signal} value with five fields: the canonical \textit{head} and \textit{tail}, and three fields used by the runtime. These runtime fields consist of a flag $U$ required by the advance semantics and two pointers $prev$ and $next$ that link the signal into the signal heap.
The \textbf{signal} constructor is evaluated like any other $\textbf{ctor}_i$ with an additional step that inserts the resulting \textbf{signal} value into the linked-list structure of the signal heap.
% When applying the \textbf{signal} constructor with a set of arguments, each argument is evaluated and stored into a fresh memory location, just as with other constructors, but the resulting \textbf{signal} value is also inserted into the signal heap.
When a signal is later erased, its fields are decremented and the signal is removed from the signal heap.

%GOAL: the delay constructor should have call-by-name semantics.
There is one additional adjustment required for \texttt{Delayed} values. In particular, the constructor $\mathsf{delay} \; e$ should behave as call-by-name, but this is not enforced by operational semantics since such values are treated as $\textbf{ctor}_i$ values. Instead the AST is rewritten so that expressions of the form $\mathsf{delay} \; e$ become $\mathsf{delay} \; (\lambda x.e$).
% \todo{This instead costs runtime memory, since we must effectively allocate a closure AND a constructor value for $\mathsf{delay}$}.

%GOAL: Where do we reference count in reactive semantics? Signals are created, data is overwritten. Immediate erasure of intermediate signals. This must be managed.
This leaves us with the interaction between reference counting and the reactive semantics.
Although we treat the advance semantics as a black-box, its interface must consume an RC token of any the later value it receives. When the update semantics call advance on a signal's tail, advance is therefore responsible for decrementing that tail once it is no longer needed.
This ensures later values, and the data they reference, are released as early as possible. This is crucial for the success of \textbf{reset}/\textbf{reuse}. If advance only borrowed the later value, the update semantics would have to perform the decrement after advance returned, which would block reuse of memory captured inside the later.
We base these requirements on the result of running the program in Appendix~\ref{appendix:advance_func} through the reference counting algorithm of Section~\ref{section:refcount} to achieve the pseudo-code from Appendix~\ref{appendix:advance_ir}.

With this interface in place, reference counting in the update semantics proceeds as: if a signal does not produce a new value, then there is nothing to do. If it does produce a value, the original head is decremented before advancing the tail, because it is stale and it may hold onto references, which could block reuse. The head and tail of the intermediate signal are incremented before being copied back into the original signal.
Once the entire heap has been updated, the value that triggered the update process is decremented to release it.

% \todo{is this a better spot to show the heap? We could show example of general heap to show how the signal heap is just a subset}

Figure~\ref{figure:semantics:heap_vis} visualises the Rizzo heap as a sequence of mappings from location $l_i$ to value and reference counter $C_i$. The heap stores regular constructor values alongside signal values. 
The signal heap is linked together by the \texttt{prev} and \texttt{next} fields.
In Figure~\ref{figure:semantics:heap_vis}, the heap base is $l_2$, whose \texttt{prev} field is empty and whose \texttt{next} field forms the signal heap $l_2, l_n, l_m$.
\todo{Make the above part flow better. I think its a fine spot to bring this up. We have introduced all the necessary information/concepts but it reads weird.}

\begin{figure}[h]
    \centering
    \[\begin{array}{rcl}
        l_0 &\to& \textbf{ctor}_0(),\ C_0 \\
        l_1 &\to& \textbf{ctor}_0(42, 12),\ C_1 \\
        l_2 &\to& \textbf{signal}(l_0, l_1),\ \texttt{prev}=\cdot,\ \texttt{next}=l_n,\ C_2 \\
        l_3 &\to& \textbf{ctor}_0(),\ C_3 \\
        l_4 &\to& \textbf{ctor}_1(5),\ C_4 \\
        l_5 &\to& \textbf{ctor}_0(l_3, l_4),\ C_5 \\
        \vdots & & \vdots \\
        l_{n-1} &\to& \textbf{ctor}_6(l_{n-2}, l_{n-3}),\ C_{n-1} \\
        l_n &\to& \textbf{signal}(\texttt{"Hello, World!"}, l_{n-1}),\ \texttt{prev}=l_5,\ \texttt{next}=l_m,\ C_n
    \end{array}\]
    \caption{The Rizzo heap storing both constructor- and signal values. The signal heap is linked together by the \texttt{prev} and \texttt{next} fields}\label{figure:semantics:heap_vis}
\end{figure}


% For RizzoLang we need to adhere to the \texttt{advance} and \texttt{update} semantics of Rizzo as defined in \citetitle{rizzo_modal_2025} \cite{rizzo_modal_2025}, but we also need to define the semantics of reference counting for our implementation. In this section we first describe the evaluation strategy of Rizzo, then the reactive semantics for signals, and finally how reference counting interacts with the heap structure.

% % Rizzo has two layers of semantics: the standard evaluation semantics for expressions, and the reactive semantics that govern how signals are updated over time. Understanding both is essential for implementing an efficient runtime.

% \subsubsection{Eager Evaluation}

% Rizzo uses eager (strict) evaluation for all expressions. Function arguments are evaluated before the function body executes, and let bindings evaluate their right-hand side before the body. This is in contrast to lazy languages like Haskell where evaluation is deferred until values are needed.

% However, there is an important exception: \texttt{Delayed} and \texttt{Later} computations are not evaluated immediately. When we write \texttt{delay e}, the expression \texttt{e} is not evaluated---instead, a thunk (suspended computation) is created. Similarly, when we write \texttt{f |> later}, the application of \texttt{f} is deferred until the \texttt{later} value becomes available.

% This combination of eager evaluation with explicit delayed computations gives programmers precise control over when computations occur. Signal combinators use \texttt{delay} to ensure that recursive calls happen only when time advances, preventing infinite loops.

% \subsubsection{Reactive Semantics: Time Steps and Updates}

% The reactive semantics describe how signals evolve over time. Time in Rizzo is discrete: it advances in steps, with each step triggered by input arriving on a channel. Between time steps, signals maintain their current values.

% When a channel produces a value, the runtime must update all signals that depend (directly or indirectly) on that channel. This is done through two key operations: \texttt{ticked} and \texttt{advance}.

% \paragraph{The Ticked Predicate.} Given a \texttt{Later} computation and a channel that just produced a value, the \texttt{ticked} predicate determines whether the computation should produce a value in this time step:

% \begin{itemize}
%     \item $\texttt{ticked}(\texttt{never}, c, v) = \texttt{false}$ --- never produces a value
%     \item $\texttt{ticked}(\texttt{wait } k, c, v) = (k = c)$ --- ticks when the waited channel fires
%     \item $\texttt{ticked}(\texttt{tail } s, c, v) = \texttt{updated}(s)$ --- ticks when the signal was updated
%     \item $\texttt{ticked}(\texttt{sync } l_1 \; l_2, c, v) = \texttt{ticked}(l_1, c, v) \lor \texttt{ticked}(l_2, c, v)$
%     \item $\texttt{ticked}(f \triangleright l, c, v) = \texttt{ticked}(l, c, v)$ --- follows the later argument
% \end{itemize}

% \paragraph{The Advance Operation.} When a \texttt{Later} computation has ticked, the \texttt{advance} operation extracts the value it produces:

% \begin{itemize}
%     \item $\texttt{advance}(\texttt{wait } k, c, v) = v$ --- returns the channel's value
%     \item $\texttt{advance}(\texttt{tail } s, c, v) = s$ --- returns the (now updated) signal
%     \item $\texttt{advance}(f \triangleright l, c, v) = f(\texttt{advance}(l, c, v))$ --- applies the delayed function
%     \item $\texttt{advance}(\texttt{sync } l_1 \; l_2, c, v)$ returns \texttt{Left}, \texttt{Right}, or \texttt{Both} depending on which computations ticked
% \end{itemize}

% \paragraph{The Heap Update Procedure.} The runtime maintains all signals in a heap (see Section~\ref{section:rizzolang:heap}). When a channel fires, the update procedure iterates through the heap:

% \begin{enumerate}
%     \item For each signal $s$ with tail $t$:
%     \item If $\texttt{ticked}(t, c, v)$:
%     \begin{enumerate}
%         \item Compute $s' = \texttt{advance}(t, c, v)$ --- this produces a new signal
%         \item Copy $\texttt{head}(s')$ and $\texttt{tail}(s')$ back into $s$
%         \item Mark $s$ as updated
%         \item The intermediate signal $s'$ can now be deallocated
%     \end{enumerate}
%     \item Otherwise, mark $s$ as not updated
% \end{enumerate}

% The copying step is crucial: it ensures that references to $s$ remain valid while the signal's contents are updated. This is the ``mutable reference'' semantics that distinguishes Rizzo from other FRP languages.
% % this will become important as to why we cant reuse the intermediate signal

% \subsubsection{Update Ordering and Cascading}

% The order in which signals are updated matters. Consider a signal $s_2$ whose tail depends on another signal $s_1$. If $s_1$ updates, then $s_2$ may also need to update (if its tail uses \texttt{tail} $s_1$). For correct semantics, $s_1$ must be updated before $s_2$.

% The heap maintains signals in creation order, which ensures that dependencies are respected: a signal can only depend on signals created before it. When a new signal is created during an update (e.g., by advancing a \texttt{Later} computation), it is inserted into the now heap, ensuring the signal will not be updated until the next time step. This allows for cascading updates without risking infinite loops within a single time step.

% \subsubsection{Reference Counting Semantics}

% The \citetitle{ullrich_counting_2021} paper defines a reference-counted intermediate representation ($\lambda_{\text{RC}}$) with explicit \texttt{inc}, \texttt{dec}, \texttt{reset}, and \texttt{reuse} operations. Our compiler transforms Rizzo programs into this representation before generating C code.

% The key semantic rules are:

% \begin{itemize}
%     \item Every variable starts with a reference count of 1 when bound
%     \item Passing a variable to an \emph{owned} parameter consumes one reference
%     \item Passing a variable to a \emph{borrowed} parameter does not consume a reference
%     \item \texttt{inc x} increments the reference count of \texttt{x}
%     \item \texttt{dec x} decrements the reference count; if it reaches zero, the object is freed
%     \item \texttt{reset x} prepares \texttt{x} for potential reuse if its count is 1
%     \item \texttt{reuse x in C(...)} reuses the memory of \texttt{x} for a new constructor
% \end{itemize}

% These operations are inserted automatically by the compiler based on ownership analysis. The details of this transformation are covered in Section~\ref{section:refcount}.